Capstone Movielens Report

### 2025-09-25 reports/capstone-movielens.site/01.intro.Rmd
Fixes for citations.

The goal of the project is to build a Recommendation System using a [10M version of the MovieLens dataset](http://grouplens.org/datasets/movielens/10m/).
Following the [Netflix Grand Prize Contest](https://archive.nytimes.com/bits.blogs.nytimes.com/2009/09/21/netflix-awards-1-million-prize-and-starts-a-new-contest/index.html) requirements, we will evaluate the _Root Mean Square Error_ (_RMSE_) score, which, as shown in [Section 23.2 Loss function](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#sec-netflix-loss-function) of the _Course Textbook_, is defined as:
$$
\mbox{RMSE} = \sqrt{\frac{1}{N} \sum_{i,j}^{N} (y_{i,j} - \hat{y}_{i,j})^2}
$$

@@

Now, note the expressions below which confirm the fact explained in [Section _23.1.1 Movielens data_](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_[@IDS2] that not every user rated every movie:
```{r}
max_possible_ratings <- n_movies*n_users
sprintf("Maximum possible ratings: %s", max_possible_ratings)
sprintf("Rows in `edx` dataset: %s", dim_edx[1])
sprintf("Not every movie was rated: %s", max_possible_ratings > dim_edx[1])

>>>>

The goal of the project is to build a Recommendation System using a [10M version of the MovieLens dataset](http://grouplens.org/datasets/movielens/10m/).
Following the [Netflix Grand Prize Contest](https://archive.nytimes.com/bits.blogs.nytimes.com/2009/09/21/netflix-awards-1-million-prize-and-starts-a-new-contest/index.html) requirements, we will evaluate the _Root Mean Square Error_ (_RMSE_) score, which, as shown in [Section 23.2 *Loss function*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#sec-netflix-loss-function) of the _Course Textbook_, is defined as:
$$
\mbox{RMSE} = \sqrt{\frac{1}{N} \sum_{i,j}^{N} (y_{i,j} - \hat{y}_{i,j})^2}
$$

@@

Now, note the expressions below which confirm the fact explained in [Section 23.1.1 *Movielens data*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_ that not every user rated every movie[@IDS2_23-1-1]:
```{r}
max_possible_ratings <- n_movies*n_users
sprintf("Maximum possible ratings: %s", max_possible_ratings)
sprintf("Rows in `edx` dataset: %s", dim_edx[1])
sprintf("Not every movie was rated: %s", max_possible_ratings > dim_edx[1])

### 2025-09-25 reports/capstone-movielens.site/02-0.methods.Rmd
Fixes for citations.

We will split the `edx` dataset into a training set, which we will use to build and train our models, and a test set in which we will compute the accuracy of our predictions, the way described in [Section 23.1.1 Movielens data](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_ mentioned above[@IDS2_23-1-1].We will also use the _5-Fold Cross Validation_ method as described in [Section 29.6 Cross validation](https://rafalab.dfci.harvard.edu/dsbook-part-2/ml/resampling-methods.html#cross-validation) of the _Course Textbook_[@IDS2_29-6]. 
To prepare datasets for processing, we will use the following functions, specifically designed for these operations:

@@

We will use the array representation described in [Section 17.5 Analysis of variance (ANOVA)](https://rafalab.dfci.harvard.edu/dsbook-part-2/linear-models/treatment-effect-models.html#sec-anova) of the _Course Textbook_[@IDS2_17-5], for the training data. For this purpose, we will convert the `edx` dataset into a matrix using the `tidyr::pivot_wider` function:

@@

Here we have a list of sample objects we need to perform the _5-Fold Cross Validation_ 
as explained in [Section 29.6.1 K-fold cross validation](https://rafalab.dfci.harvard.edu/dsbook-part-2/ml/resampling-methods.html#k-fold-cross-validation) of the _Course Textbook_[@IDS2_29-6-1]:
```{r eval=FALSE}
  start <- put_start_date()
  edx_CV <- lapply(kfold_index,  function(fold_i){

@@

As explained in [Section  23.4 _User effects_](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#user-effects) of the _Course Textbook_[@IDS2_23-4] we know ratings can’t be below 0.5 or above 5. For this reason, we will use the `clamp` function described in that section:
```{r eval=FALSE}
clamp <- function(x, min = 0.5, max = 5) pmax(pmin(x, max), min)
```

>>>>

We will split the `edx` dataset into a training set, which we will use to build and train our models, and a test set in which we will compute the accuracy of our predictions, the way described in [Section 23.1.1 *Movielens data*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_ mentioned above[@IDS2_23-1-1].We will also use the _5-Fold Cross Validation_ method as described in [Section 29.6 *Cross validation*](https://rafalab.dfci.harvard.edu/dsbook-part-2/ml/resampling-methods.html#cross-validation) of the _Course Textbook_[@IDS2_29-6]. 
To prepare datasets for processing, we will use the following functions, specifically designed for these operations:

@@

We will use the array representation described in [Section 17.5 *Analysis of variance (ANOVA)*](https://rafalab.dfci.harvard.edu/dsbook-part-2/linear-models/treatment-effect-models.html#sec-anova) of the _Course Textbook_, for the training data[@IDS2_17-5]. For this purpose, we will convert the `edx` dataset into a matrix using the `tidyr::pivot_wider` function:

@@

Here we have a list of sample objects we need to perform the _5-Fold Cross Validation_ 
as explained in [Section 29.6.1 *K-fold cross validation*](https://rafalab.dfci.harvard.edu/dsbook-part-2/ml/resampling-methods.html#k-fold-cross-validation) of the _Course Textbook_[@IDS2_29-6-1]:
```{r eval=FALSE}
  start <- put_start_date()
  edx_CV <- lapply(kfold_index,  function(fold_i){

@@

As explained in [Section  23.4 *User effects*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#user-effects) of the _Course Textbook_ we know ratings cannot be below 0.5 or above 5. For this reason, we will use the `clamp` function described in that section[@IDS2_23-4]:
```{r eval=FALSE}
clamp <- function(x, min = 0.5, max = 5) pmax(pmin(x, max), min)
```

### 2025-09-25 reports/capstone-movielens.site/02-3.OMR-model.Rmd
Minor fix for sections  1-2.3.
Fixes for citations.

Let's begin our analysis by evaluating the simplest model described in [Section 23.3 _A first model_](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#a-first-model) of the *Course Textbook*[@IDS2_23-3], and then gradually refine it through further research.

@@

with $\varepsilon_{i,j}$ independent errors sampled from the same distribution centered at 0 and $\mu$ is the rating for all movies.

>>>>

Let's begin our analysis by evaluating the simplest model described in [Section 23.3 *A first model*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#a-first-model) of the *Course Textbook*[@IDS2_23-3], and then gradually refine it through further research.

@@

with $\varepsilon_{i,j}$ independent errors sampled from the same distribution centered at 0 and $\mu$ is the average rating for all movies.

### 2025-09-25 reports/capstone-movielens.site/02.6.UMGE-model.Rmd
Fixes for citations.

As mentioned in [Section 23.7: Exercises](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#exercises) of the _Chapter "23 Regularization" of the Course Textbook_ the `Movielens` dataset also has a genres column. This column includes every genre that applies to the movie (some movies fall under several genres)[@IDS2_23-7].

@@

To account for a _genre effect_ we will use the model suggested in the [Section 23.7: Exercises](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#exercises) of the _Chapter 23 "Regularization" of the Course Textbook_[@IDS2_23-7]:

If we define a _genre treatment effect_ $g_{i,j}$ for user's $i$ rating of movie $j$, we can use the following models to account for the `genre` effect:

>>>>

As mentioned in [Chapter 23: *Regularization*, Section 23.7: *Exercises*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#exercises) of the *Course Textbook* the `Movielens` dataset also has a genres column. This column includes every genre that applies to the movie (some movies fall under several genres)[@IDS2_23-7].

@@

To account for a _genre effect_ we will use the model suggested in the [Chapter 23: *Regularization*, Section 23.7: *Exercises*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#exercises) of the *Course Textbook*:

If we define a _genre treatment effect_ $g_{i,j}$ for user's $i$ rating of movie $j$, we can use the following models to account for the `genre` effect[@IDS2_23-7]:
